Linear stability of magnetized massive protoplanetary disks

### 2014-05-28 paper/linear.tex
first draft, spell-checked
revision to 1st ref report

\section{Linear problem}\label{linear}
We consider axisymmetric Eulerian perturbations to the above
equilibrium in the form $\real[\dd\rho(z)\exp{\imgi(k_xx + \sigma
    t)}]$ and similarly for other fluid variables. Here, $k_x$ is a
constant radial wavenumber and $\sigma=-(\omega +\imgi\gamma)$ is a
complex  frequency, where $-\omega$ is the real mode frequency and $\gamma$ is
the growth rate. We take $k_x>0$ without loss of generality. 
Hereafter, we surpress the exponential factor and the real part notation.

@@

\subsection{Domain and boundary conditions}\label{domain}

\subsubsection{Case 1: vertical field}
We impose $\bm{U}$ to be an even function of $z$, so that
$d\bm{U}/dz=0$ at $z=0$. This permits higher numerical resolution by
reducing the computational domain to $z\in~[0,Z_s]$.  
At the upper disk boundary $z=Z_s$ we set 
\begin{align}
  \dbx(Z_s) = \dby(Z_s) = \dvz (Z_s) = 0, 
\end{align}
so the field remains vertical. The potential boundary
condition is 
\begin{align}
  \dphi^\prime(\zmax) + k_x\dphi(\zmax) = 0 \label{pot_bc1},
\end{align}
and is adapted from \cite{goldreich65a}. The general form of this
potential boundary condition is used for case 2 below.

\subsubsection{Case 2: tilted field} 
In this more general setup the computational domain is
$z\in[-Z_s, Z_s]$. At the disk surfaces we adopt the `halo' model of
\cite{gammie94}, so that
\begin{align}
  \Delta(\pm Z_s) & = 0, \\
  \dby(\pm Z_s) & = 0,\\
  \dbz(\pm Z_s) \mp \imgi \dbx(\pm Z_s) &= 0,
\end{align}
and the potential boundary condition is 
\begin{align}
  \dphi^\prime(\pm \zmax) \pm k_x\dphi(\pm\zmax) =  - \left. \frac{\Omega^2
    \rho \xi_z}{\rho_0Q}\right|_{\pm\zmax}, \label{pot_bc2}
\end{align}
where $\xi_z=\dvz/\imgi\sigma$ is the vertical Lagrangian
displacement. In practice,  Eq. \ref{pot_bc1} and \ref{pot_bc2} are
very similar because $\rho(\pm\zmax)\ll\rho_0$.

@@

The above procedure discretizes the linear equations to a matrix equation,
\begin{align}\label{matrix_eqn}
\bm{M}\bm{w} = \mathbf{0}, 
\end{align}
where $\bm{M}$ is a $4N_z\times 4 N_z$ matrix representing the $L_{ij}$ 
plus boundary conditions, 
and $\bm{w}$ is a vector storing the pseudo-spectral coefficients. 
Starting with an initial guess for $\sigma$, non-trivial solutions to
Eq. \ref{matrix_eqn} are obtained by varying $\sigma$ using Newton-Raphson iteration 
such that $\mathrm{det}\bm{M}=0$ \citep[details can be found in][]{lin12}.

>>>>

\section{Linear problem}\label{linear}
We consider axisymmetric Eulerian perturbations to the above
equilibrium in the form $\real[\dd\rho(z)\exp{\imgi(k_xx + \sigma
    t)}]$ and similarly for other fluid variables. Here, $k_x$ is a
constant radial wavenumber and $\sigma=-(\omega +\imgi\gamma)$ is a
complex  frequency, where $-\omega$ is the real mode frequency and $\gamma$ is
the growth rate. We take $k_x>0$ without loss of generality. 
Hereafter, we suppress the exponential factor and the real part notation.

@@

\subsection{Domain and boundary conditions}\label{domain}
{\bf 
For a vertical field, considered in \S\ref{result1} and
\S\ref{result2}, we take $\bm{U}$ to be an even function of
$z$. Odd modes are permitted in \S\ref{result3}, where an azimuthal field may
be included. In both
setups the gravitational potential boundary condition, given by
\cite{goldreich65a}, is
%The potential boundary
%condition is 
%\begin{align}
%  \dphi^\prime(\zmax) + k_x\dphi(\zmax) = 0 \label{pot_bc1},
%\end{align}
%and is adapted from \cite{goldreich65a}. The general form of this
%potential boundary condition is used for case 2 below.
%and 
%the potential boundary condition is 
\begin{align}
  \dphi^\prime(\pm \zmax) \pm k_x\dphi(\pm\zmax) =  - \left. \frac{\Omega^2
    \rho \xi_z}{\rho_0Q}\right|_{\pm\zmax}, \label{pot_bc2}
\end{align}
where $\xi_z=\dvz/\imgi\sigma$ is the vertical Lagrangian
displacement, and $z=\pm Z_s$ is the upper and lower disk surfaces,
respectively. 
}
%In practice,  Eq. \ref{pot_bc1} and \ref{pot_bc2} are
%very similar because $\rho(\pm\zmax)\ll\rho_0$. 
%}

\subsubsection{Case 1: vertical field}
{\bf Here we impose} $d\bm{U}/dz=0$ at $z=0$. This permits higher
numerical resolution by reducing the computational domain to
$z\in~[0,Z_s]$. At the upper disk boundary $z=Z_s$ we set  
\begin{align}
  \dbx(Z_s) = \dby(Z_s) = \dvz (Z_s) = 0, 
\end{align}
so the field remains vertical. {\bf The derivation of the magnetic
  field boundary conditions may be found in \cite{sano99}.}

\subsubsection{Case 2: tilted field} 
In this more general setup the computational domain is
$z\in[-Z_s, Z_s]$ {\bf and no symmetry across the midplane is
  enforced.} At the disk surfaces we adopt the `halo' model of
\cite{gammie94}, so that
\begin{align}
  \Delta(\pm Z_s) & = 0, \\
  \dby(\pm Z_s) & = 0,\\
  \dbz(\pm Z_s) \mp \imgi \dbx(\pm Z_s) &= 0,
\end{align}
    {\bf and this case permits $\dd v_z(\pm Z_s)\neq0$.}
%      In practice, the gravitational potential boundary condition is
%      similar to case 1 because $\rho(\pm Z_s)\ll\rho_0$.}

@@

The above procedure discretize the linear equations to a matrix equation,
\begin{align}\label{matrix_eqn}
\bm{M}\bm{w} = \mathbf{0}, 
\end{align}
where $\bm{M}$ is a $4N_z\times 4 N_z$ matrix representing the $L_{ij}$ 
plus boundary conditions, 
and $\bm{w}$ is a vector storing the pseudo-spectral coefficients. 
Starting with an initial guess for $\sigma$, non-trivial solutions to
Eq. \ref{matrix_eqn} are obtained by varying $\sigma$ using Newton-Raphson iteration 
such that $\mathrm{det}\bm{M}=0$ \citep[details can be found in][]{lin12}.

### 2014-05-28 paper/paper.tex
revision to 1st ref report

\begin{abstract}
  Magneto-rotational instability (MRI) and gravitational instability
  (GI) are the two principle routes to turbulent angular momentum
  transport in accretion disks. Protoplanetary disks may develop 
  both. This paper aims to reinvigorate interest in the study of
  magnetized  massive protoplanetary disks, starting from the basic
  issue of stability. The local linear stability of a 
  self-gravitating, uniformly magnetized, differentially rotating, 
  three-dimensional disk subject to axisymmetric perturbations is
  calculated numerically. The formulation includes 
  resistivity. It is found that the reduction in the disk thickness
  by self-gravity can decrease MRI growth rates; the MRI
  becomes global in the vertical direction, and MRI modes 
  with small radial length scales are stabilized. In massive
  disks with layered resistivity, the MRI is not well-localized
  to regions where the Elsasser number exceeds unity. For MRI modes
  with radial length scales on the order of the disk thickness, self-gravity 
  can enhance density perturbations, an effect that becomes 
  significant in the presence of a strong toroidal field. In
  gravitationally unstable disks where GI and MRI growth rates are
  comparable, the character of unstable modes can transition smoothly
  between MRI and GI. Implications for non-linear simulations are
  discussed briefly.

@@

\appendix
\input{appendix}

>>>>

\begin{abstract}
  Magneto-rotational instability (MRI) and gravitational instability
  (GI) are the two principle routes to turbulent angular momentum
  transport in accretion disks. Protoplanetary disks may develop 
  both. This paper aims to reinvigorate interest in the study of
  magnetized  massive protoplanetary disks, starting from the basic
  issue of stability. The local linear stability of a 
  self-gravitating, uniformly magnetized, differentially rotating, 
  three-dimensional disk subject to axisymmetric perturbations is
  calculated numerically. The formulation includes 
  resistivity. It is found that the reduction in the disk thickness
  by self-gravity can decrease MRI growth rates; the MRI
  becomes global in the vertical direction, and MRI modes 
  with small radial length scales are stabilized. {\bf The
    upper limit to the vertical field strength to permit the MRI
    in a strongly self-gravitating disk is estimated.} In massive
  disks with layered resistivity, the MRI is not well-localized
  to regions where the Elsasser number exceeds unity. For MRI modes
  with radial length scales on the order of the disk thickness, self-gravity 
  can enhance density perturbations, an effect that becomes 
  significant in the presence of a strong toroidal field, {\bf and
    which depends on the symmetry of the underlying MRI mode.} In
  gravitationally unstable disks where GI and MRI growth rates are
  comparable, the character of unstable modes can transition smoothly
  between MRI and GI. Implications for non-linear simulations are
  discussed briefly.

@@

\acknowledgements
{\bf I thank K. Menou, S. Fromang and A. Youdin for helpful discussions
during the course of this project. 
}
\appendix
\input{appendix}

### 2014-05-28 paper/summary.tex
revision to 1st ref report

For a vertical field, the requirement for the MRI is that   
its vertical wavelength $\lambda \lesssim H$. While $\lambda$ is independent of
the strength of self-gravity $Q$, the disk thickness $H=H(Q)$ 
decreases with increasing SG, which can reduce MRI growth rates. 
Thus, a sufficiently massive disk can potentially suppress the MRI. 
The MRI is also restricted to larger radial
scales as $Q$ is lowered. This means that the MRI becomes more global
in self-gravitating disks. Finally, the condition $\lambda < H$ 
requires $f(Q)/\sqrt{\beta}$ to be sufficiently small, so weaker
fields are required to permit the MRI with increasing self-gravity.   
%since MRI is 3D, strong SG makes it 2D

>>>>

For a vertical field, the requirement for the MRI {\bf to operate} is that   
its vertical wavelength $\lambda \lesssim 2H$. 
{\bf 
%While $\lambda$ is independent of
%the strength of self-gravity $Q$ {\bf for fixed $\beta$}, 
The disk thickness $H=H(Q)$ decreases with increasing SG. This can
reduce MRI growth rates when 
$\beta$, and hence $\lambda$, is fixed.  
}
Thus, a sufficiently massive disk can potentially suppress the MRI. 
The MRI is also restricted to larger radial 
scales as $Q$ is lowered. This means that the MRI becomes more global
in self-gravitating disks.

{\bf
  The condition $\lambda < 2H$ 
  %requires $f(Q)/\sqrt{\beta}$ to
  %  be sufficiently small (see
  %  Eq. \ref{lambda_ideal}---\ref{lambda_resis}), so larger values of
  %  $\beta$ are required to  
  %  permit the MRI with increasing self-gravity, since $f(Q)$ increases
  %  with decreasing $Q$. 
  may be written more precisely as 
  \begin{align}\label{cond1}
    \frac{n}{\mathrm{min}(\Lambda_0, 1)}\frac{f(Q)}{\sqrt{\beta}} \lesssim 1,
  \end{align}
  where $n\sim 3$ is a numerical factor and
  $\mathrm{min}(\Lambda_0,1)$ accounts for the ideal and resistive
  limits (see Eq. \ref{lambda_ideal}---\ref{lambda_resis}). Since $f$
  increases with decreasing $Q$, Eq. \ref{cond1} implies
  the MRI requires larger $\beta$ values with  
  increasing self-gravity. 
  For definiteness, consider the ideal polytropic disk. Then Eq. \ref{cond1} is
  \begin{align}\label{cond2}
    \beta^{-1/2} \lesssim
    \frac{\sqrt{15}}{4\pi}\sqrt{Q}\arccos{\left(\frac{Q}{1+Q}\right)}. 
  \end{align}
  For a non-self-gravitating disk, $Q\to\infty$ and Eq. \ref{cond2} is
  $\beta \gtrsim 16\pi^2/30\simeq 5$. For $Q\ll 1$, the condition is
  $\beta \gtrsim 64/15Q$, giving $\beta\gtrsim 20$ for
  $Q=0.2$. We confirmed this numerically, finding that the growth rate
  $\gamma  \lesssim 0.1$ when $\beta \lesssim 3.3$ for $Q=20$ and
  $\beta \lesssim 17$ for $Q=0.2$.  
%  Inserting $Q\sim 0.2$ for an ideal disk marginally stable to
%  self-gravity, we find $v_{A0}\lesssim 0.2\csmid$ for the MRI.

  We can also place an upper bound on the field strength
  $B_z$. Writing $v_{A0} = B_z\sqrt{4\pi G Q/\mu_0\Omega^2}$, we find 
  Eq. \ref{cond2} is independent of $Q$ for $Q\ll1$, and 
  \begin{align}\label{cond3}
    \frac{B_z}{\csmid\Omega}\sqrt{\frac{\pi G}{\mu_0}}\lesssim
    \frac{\sqrt{15}}{16}
  \end{align}
  is needed for the MRI to operate in an ideal polytropic disk with strong
  self-gravity. Although both the MRI wavelength and disk thickness
  vanish as $Q\to 0$, the MRI can still operate provided the field is
  sufficiently weak according to Eq. \ref{cond3}.

%Q->0, thickness -> 0

%This condition translates to an upper
%  limit on the actual field strength $B_z$. 
%
%  For definiteness, consider
%  the polytropic disk with $Q \ll 1$. The disk thickness is then 
%  $H\sim \csmid\pi\sqrt{Q}/2\Omega$. This applies at marginal
%  gravitatational stability for which $Q\sim 0.2$. Next, expressing
%  $\rho_0$ in terms of $Q$, we have $\lambda\sim 4\pi B_z \sqrt{\pi
%    GQ/\mu_0}/\Omega^2$. Then 
%  \begin{align}
%    \frac{B_z}{\csmid\Omega}\sqrt{\frac{\pi G}{\mu_0}}\lesssim
%    \frac{1}{4}
%  \end{align}
%  is needed for the MRI when self-gravity is strong. 
  % We can quanitfy this by comparing
  %$\lambda$ to the Jeans length $\lambda_J \equiv 2
  %c_s^2/G\Sigma\sim 2\csmid^2/G\rho_0 H \sim 8\pi f^2Q H$. Then $\lambda
  %< H $ translates to
  %\begin{align}    
  %\frac{\lambda}{\lambda_J}\lesssim \frac{1}{8\pi f^2(Q) Q}.
  %\end{align}
  %Given a mid-plane density $\rho_0$, and hence $Q$, the left-hand-side
  %is a measure of the field strength.

}
